---
title: "Retraction Pollution Study Report"
format:
  html:
    toc: true
    theme: cosmo
params:
  summary: ""
  crawl_summary: ""
  depth3_truncated: "false"
  pipeline_mode: "unknown"
  rw_snapshot_date: "unknown"
  oc_access_date: "unknown"
  openalex_access_date: "unknown"
  top_seeds: "data.frame()"
  depth_counts: "data.frame()"
  validation: "data.frame()"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)

fmt_kable <- function(df) {
  if (is.data.frame(df) && nrow(df) > 0L) {
    kable(df, format = "html") |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = FALSE)
  } else {
    "_(no rows)_"
  }
}

parse_summary <- function(s) {
  if (is.null(s) || !nzchar(s)) return(list())
  tryCatch(
    jsonlite::fromJSON(s, simplifyDataFrame = FALSE),
    error = function(e) {
      # Try "metric,value" CSV-shaped text
      out <- list()
      for (line in strsplit(s, "\n")[[1]]) {
        parts <- strsplit(line, ",", fixed = TRUE)[[1]]
        if (length(parts) >= 2L) out[[parts[1]]] <- parts[2]
      }
      out
    }
  )
}
```

# Provenance

This report describes the citation-pollution network constructed around a set
of retracted or concerning papers. Source snapshots were captured on the dates
below; all metrics reflect the state of each source at those points in time.

```{r provenance}
prov <- data.frame(
  Source = c("Retraction Watch (Retraction + Expression of Concern)",
             "OpenCitations (DOI-based citation index)",
             "OpenAlex (work resolution + recursive expansion)"),
  `Snapshot / Access date` = c(
    params$rw_snapshot_date,
    params$oc_access_date,
    params$openalex_access_date
  ),
  check.names = FALSE
)
fmt_kable(prov)
```

**Generated:** `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`

# Summary Metrics

```{r summary}
summary_list <- parse_summary(params$summary)
if (length(summary_list) > 0L) {
  metric_names <- names(summary_list)
  if (is.null(metric_names) || length(metric_names) != length(summary_list)) {
    metric_names <- paste0("metric_", seq_along(summary_list))
  }
  values <- vapply(summary_list, function(value) {
    if (is.null(value) || length(value) == 0L) return(NA_character_)
    as.character(value[[1]])
  }, character(1))
  df <- data.frame(
    Metric = metric_names,
    Value = values
  )
  fmt_kable(df)
} else {
  "No analysis summary found. Run `run_analysis()` first."
}
```

# Data Sources

- **Retraction Watch.** Seed records are drawn from the Retraction Watch
  database, filtered to `Retraction` and `Expression of Concern` notice
  types. Each seed carries the original DOI, notice date, and (where
  available) the OpenAlex ID of the retracted/concerning paper.
- **OpenCitations.** DOI-to-DOI citation edges are first sourced from
  OpenCitations (COCI) for every parent node that carries a DOI, at each
  frontier depth. This gives the most complete coverage of
  non-OpenAlex-indexed citations.
```{r openalex-source-note, results='asis'}
summary_list <- parse_summary(params$summary)
openalex_edges <- suppressWarnings(as.integer(summary_list[["openalex_edges"]]))
if (identical(params$pipeline_mode, "opencitations") ||
    (!is.na(openalex_edges) && openalex_edges == 0L)) {
  cat("- **OpenAlex.** Not used for this OpenCitations-only output; OpenAlex edge count is 0.\n")
} else {
  cat("- **OpenAlex.** Used to resolve seeds and frontier nodes to canonical work records and to perform recursive referenced-works expansion when OpenCitations returns no edges.\n")
}
```

# Network Size by Depth

```{r incomplete-crawl-warning, results='asis'}
if (is.data.frame(params$validation) &&
    any(params$validation$check == "frontier_processed_for_depth2_claim" &
        params$validation$status == "fail")) {
  cat("**Incomplete crawl warning:** depth-2 graph counts are lower-bound counts from the processed frontier, not complete depth-2 census values.\n\n")
}
```

```{r depth-counts-fig}
depth_png <- "figures/depth_counts.png"
if (file.exists(depth_png)) {
  knitr::include_graphics(depth_png)
} else if (is.data.frame(params$depth_counts) && nrow(params$depth_counts) > 0L) {
  library(ggplot2)
  ggplot(params$depth_counts,
         aes(x = factor(.data$depth), y = .data$node_count)) +
    geom_col(fill = "#3b82f6") +
    labs(x = "Network depth", y = "Nodes",
         title = "Citation Pollution Network Size by Depth") +
    theme_minimal()
} else {
  "_(depth_counts figure and data not available)_"
}
```

```{r depth-counts-table}
if (is.data.frame(params$depth_counts) && nrow(params$depth_counts) > 0L) {
  fmt_kable(params$depth_counts)
}
```

# Most Polluted Seeds

The following retracted or concerning papers have the largest unique
descendant reach through depth 2 (direct citers plus their citers).

```{r top-seeds-table}
if (is.data.frame(params$top_seeds) && nrow(params$top_seeds) > 0L) {
  show_cols <- intersect(
    c("record_id", "title", "notice_type", "direct_citers",
      "seed_record_count", "depth2_descendants", "total_depth2_reach",
      "post_notice_direct_citers", "post_notice_depth2_descendants"),
    names(params$top_seeds)
  )
  fmt_kable(utils::head(params$top_seeds[, show_cols, drop = FALSE], 10L))
} else {
  "_(no seed metrics available)_"
}
```

```{r top-seeds-fig}
top_png <- "figures/top_polluted_seeds.png"
if (file.exists(top_png)) {
  knitr::include_graphics(top_png)
}
```

# Post-Notice Citation Timeline

```{r post-notice-fig, results='asis'}
timeline_png <- "figures/post_notice_timeline.png"
if (file.exists(timeline_png)) {
  knitr::include_graphics(timeline_png)
  cat("\n\nThe chart shows direct citations to seed papers whose publication date falls **after** the seed's retraction / EOC notice date. This is a descriptive count, not a citation-rate model or causal effect of the notice on later citation behavior.\n")
} else {
  "_(No post-notice citations were found; either no seed has a parseable notice date, or no citing work was published after its target's notice date.)_"
}
```

# Artifact Validation

```{r validation}
if (is.data.frame(params$validation) && nrow(params$validation) > 0L) {
  fmt_kable(params$validation)
} else {
  "_(no artifact validation table available)_"
}
```

# Collection Notes

- **Depth-2 scope.** The headline tables export the induced frontier graph
  through depth 2 (seeds -> direct citers -> depth-2 descendants). The artifact
  validation table records whether the crawl state supports a complete-depth
  claim; when that check fails, depth-2 metrics are lower bounds from the
  processed frontier.
- **Depth-3 cap.** Depth-3 expansion is optional and capped to keep
  runtime bounded. The `depth3_truncated` flag below records whether the
  depth-3 frontier was cut short.
```{r collection-mode-note, results='asis'}
if (identical(params$pipeline_mode, "opencitations")) {
  cat("- **Collection mode.** This output was generated in OpenCitations-only mode; OpenAlex metadata enrichment and fallback expansion are not present.\n")
} else {
  cat("- **OpenCitations-first ordering.** At each frontier depth, OpenCitations is queried first for DOI-bearing parent nodes; OpenAlex fills gaps and provides recursive expansion where OpenCitations returns nothing.\n")
}
```
- **Min-depth edge merge.** Citation edges are stored with a min-depth
  merge: when the same (source, target) edge is observed at multiple
  depths, the shallowest depth wins. This preserves the earliest
  path length at which a citation relationship was discovered.

**Depth-3 truncation flag:** `r params$depth3_truncated`

# Crawl Metadata

```{r crawl-summary}
cat("```json\n")
cat(params$crawl_summary)
cat("\n```\n")
```
